---
title: "Introduction to Programming with R"
subtitle: "Basic Syntax and Operators"
author: "Selene R. Schmittling"
date: "2025-08-01" 
output: 
  pdf_document:
  html_document: 
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# RStudio Environment


```{r Plot_tab_example}
# Lets create a little plot
# create some data
data <- data.frame(a = c(72,41,54,36), b=c('East','West','North','South'))

# generate plot
barplot(data$a, 
        names.arg = data$b, 
        col="blue", 
        ylab="# of Regional Managers")

# function name
# function parameters
# return values

```


# Variables in R

## Will it Variable?

```{r Variables}

wolves  <-  5
rabbits <-  20

pctWolfStudents <- .30
pctRabbitStudents <- .05

wolf1.name <- "Scruffy"
wolf2.name <- "Belinda"

wolf1.bday <- "8/18/2025"
wolf2.bday <- as.POSIXct("8/18/2025", format="%m/%d/%Y")

wolf1.isSmiley <- TRUE
wolf2.isSmiley <- FALSE

# See the variable value
wolves
wolf1.name
wolf2.bday
wolf1.isSmiley
```


# Data Types

## Numeric

```{r Integers}

wolves  <-  5
rabbits <-  20
wolves
rabbits

# Uncomment the line below.
# Look for a red x to the left of the line number
# Try to run the line of code.

# wolfStudentsInUS <- 13,433

# Remember to re-comment the line because the error may keep the notebook from rendering.
```

> REMEMBER: computers don't need commas in numbers. That's just to make them easier to read for us humans.

```{r Doubles-no-not-tennis}

# Now we'll define some floating point variables

pctWolfStudents <- .30
pctRabbitStudents <- .05

typeof(wolves)
typeof(pctWolfStudents)
# stillNoCommas <- 4,325.98

NCwolves <- 1098.0
# will print as an integer if the value to the right of the decimal = 0
NCwolves

SCwolves <- 1329.9
SCwolves
```


## Text

```{r}
# Some text examples

wolf1.name <- "Scruffy"
wolf2.name <- "Belinda"

wolf1.slogan <- "Howl like its 1999!!"
wolf1.FavoriteArtist <- "Prince"

wolfQuote1 <- 'He had learned well the law of club and fang.'

# What if you have quoted text in your string?
wolfQuote2 <- "'You poor devil,' said John Thornton, and Buck licked his hand."
print(wolfQuote1)
print(wolfQuote2)

# Is this ok?
wolfQuote3 <- '"You poor devil," said John Thornton, and Buck licked his hand.'
print(wolfQuote3)

```

## Dates & Times

```{r Text-Dates}
aDateButItsReallyAString <- "07/15/2025"
print(aDateButItsReallyAString)

```


```{r Text-Dates-Are-Bad-At-Math}
# To explore when data strings break, uncomment the next line and try to run it. We are just trying to find out what the date is 4 days from our date string:

# letsAdd4Days <- aDateButItsReallyAString + 4

# !! Recomment it when you are done. Otherwise you notebook won't render.
```


```{r Date-Dates}
# Let's try some dates

wolf1.bday.text <- "8/18/2019"
print(wolf1.bday.text)

wolf1.bday <- as.Date(wolf1.bday.text, format = "%m/%d/%Y")
print(wolf1.bday)

wolf2.bday <- as.POSIXct("11-01-2015", format="%m-%d-%Y")
print(wolf2.bday)
```


```{r date-format}

print(format(wolf2.bday, "%A, %B %d, %Y"))

```


```{r Date-Dates-Math}

# wolf1.bday.plus20 <- wolf1.bday.text + 20

wolf1.bday.plus20 <- wolf1.bday + 20
print(wolf1.bday.plus20)

```

## Logical

```{r Boolean-Examples}
# Example of boolean variables
isWolf1 = TRUE
isWolf2 = FALSE # coyote!!! oops!
isMajestic = TRUE
isAlive = FALSE


```

# Data Structures

## Vectors (1-D Arrays)

### How to Create Vectors

```{r Define-Vectors}
# Let's clean up the variables to make it easier to see what's going on
rm(list=ls())

# EXAMPLES OF HOW TO CREATE VECTORS
# (Single Data Type)

# Integer. Ok, fine, double.
# What if we want to store the number of NCSU wolves prowling
# around 5 different locations on campus?
numWolves = c(10, 4, 7, 12, 3) 
print(numWolves)

# Float. Ok, fine, double
# We want to store the weights of 6 wolves prowling around campus
wtWolves = c(46.8, 75.1, 62.9, 51.7, 49.0, 70.8)
print(wtWolves)

# Text.
# We want to reference the five locations at which we counted wolves
locWolves = c("Talley", "SAS Hall", "Hunt Library", "Hill Library", "Burlington Lab")
print(locWolves)

# Boolean/Logical
# We want to know if the wolves we weighed are female
isFemaleWolf = c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)
print(isFemaleWolf)

# Vectors only store one data type...really!
# We decide inappropriately to store a bunch a information about
# our favorite Wolf in a vector
# We store: name, weight, isFemale, location and age
badWolfVector <- c("Basil", 60.1, TRUE, "SAS Hall", 3) # so far so good
print(badWolfVector) # oops...

```

### Accessing Elements in a Vector

```{r Access-Elements-In-Vectors}
# Make sure you ran the code above to define the vectors before accessing them. 
# Click on the green run button in the code chunk above.

# VECTORS

# Single index
# grab the 3rd count from numWolves (# of wolves at Hunt Library)
numHuntWolves <- numWolves[3]
print(numHuntWolves)

# Range of values
# Let's grab the 2-4 wts of our wolves
subsetWts <- wtWolves[2:4]

# list of indices
idxOI <- c(1, 3, 5)
print(locWolves[idxOI])

# Out of range
# What happens if we do the following?
print(isFemaleWolf[7]) # can't access what you haven't defined


```

## Matrices (or n-D Arrays)

```{r Two-Dimension-Matrices}
# Create a 2-D matrix
# We need to store counts of Wolves in 5 locations over the course of 7 days
# a 2-D matrix would work well: one data type and we can have the rows represent
# locations and the columns = days
data <- floor(runif(35, min=0, max=50)) # generates some random data
 
wolfCounts <-  matrix(data, nrow=5, ncol=7, byrow = FALSE)

# Define Days of Week so we can use them again
daysOfWeek <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
                         "Saturday", "Sunday")
# for ease of use, we can set rownames and colnames to be descriptive
rownames(wolfCounts) = locWolves
colnames(wolfCounts) = daysOfWeek
print(wolfCounts)


```


```{r Multi-dim-matrices}
# Create a 3-D matrix
# We now want to store counts of wolves at 5 locations, 7 days a week, for 4
# weeks. 
data <- floor(runif(140, min=0, max=50)) # generates more uniform data

locWolfCounts <- array(data, dim = c(5, 7, 4),
                       dimnames = list(locWolves, daysOfWeek, 
                                       c("Week1", "Week2", "Week3", "Week4")))
print(locWolfCounts)

```

### Accessing Elements in a Matrix

```{r access-matrix}
# Make sure you run the code above to create the matrices wolfCounts and 
# locWolfCounts

# Accessing an element in a 2-D matrix requires two indices: one for row and
# one for column

# Extract the number of wolves in Talley on Wednesday
wolfCounts[1,3]
# If you provided row and column names, you can use those
wolfCounts["Talley", "Wednesday"]

# What if we want to see the counts for Talley on all days?
wolfCounts[1,]
# using row name
wolfCounts["Talley", ]

# How about if we want to see the counts for all locations on Wednesday?
wolfCounts[,3]
wolfCounts[,"Wednesday"]

# To access data in a multi-dimensional array you just need more indices
# We can access the wolfCounts at Burlington Lab on Tuesday of the 3rd week
locWolfCounts
locWolfCounts[5, 2, 3] # row, column, dimension

# We can also use row, column and dimension names
print(locWolfCounts["Burlington Lab", "Tuesday", "Week3"])

# We can set the value of data using indices too
wolfCounts
wolfCounts["Hill Library", "Friday"] <- 33
wolfCounts

```

## Dataframes (Tabular Data)

```{r Data-frames}
# We want to store data about some of our favorite Wolves who hang out at
# NC State.

wolfData <- data.frame(Name = c("Snoopy", "Fang", "Wolfie", "Artemis", 
                                "Fen", "Aurora"),
                       isFemale = c(FALSE, TRUE, FALSE, TRUE, FALSE, TRUE),
                       weight = c(62.9, 70.1, 79.2, 50.5, 68.3, 49.7),
                       favHangout = c("Talley", "Burlington Lab", 
                                      "Hunt Library", "SAS Hall", "Talley",
                                      "Hill Library"))
print(wolfData)

# Extract data from data frames
# We use indices like we do with vectors and matrices to extract or set data

# Extract Column data
# Extract the names of all the wolves
wolfData["Name"] 
# Extract data in first two columns
wolfData[c("Name","isFemale")] 
wolfData[1:2]

# Extract Data from non-contiguous columns
wolfData[c("Name", "favHangout")]
wolfData[c(1,4)]

# Extract row data
# Extract data for wolf in row 1
wolfData[1, ]

# Extract data for wolves in rows 1-3
wolfData[1:3, ]

# Extract data for wolves in rows 2 and 5
wolfData[c(2,5), ]

# Extract data for female wolves only
wolfData[wolfData$isFemale, ]
```

# Operators

## Arithmetic

```{r Unary-Binary-Operators}
# Let's clear out our variables
rm(list=ls())

# Example of unary -
numWolves = 5
reduceWolves <- -numWolves
print(reduceWolves)
  
# Example of binary -
babyWolves <- 5
totalWolves <- 10
adultWolves <- totalWolves - babyWolves
print(adultWolves)

```


```{r Arithmetic-Operators-1}
# Use operators with values
5 + 6
5 - 6
5 * 6
5 / 6
5 ^ 6
5 ** 6

# Separate Output
print('* * * * * * * * * * *')

# Use familiar operators with variables
# We'll be creative and use new values


# Apply operators to variables

adultWolves + babyWolves
adultWolves - babyWolves
adultWolves * babyWolves
adultWolves / babyWolves
adultWolves ^ babyWolves
adultWolves ** babyWolves


```

```{r Arithmetic-Operators-2}
# We want to know how many treats we have for each wolf
numWolfTreats <- 10
numWolvesTalley <- 2 # divides evenly
numWolvesHunt <- 3 # doesn't divide evenly

# Integer Division & Modulus numWolvesTalley
numWolfTreats %/% numWolvesTalley
numWolfTreats %% numWolvesTalley

# So, wolves at valley get 5 treats and there are no leftovers

# Separate Output
print("* * * * * * * * * *")

# Integer Division & Modulus using numWolvesHunt
numWolfTreats %/% numWolvesHunt
numWolfTreats %% numWolvesHunt

# Wolves at Hunt, on the other hand get three treats and there is 1 left over.
# Good luck with that.

```

## Relational

```{r Relational-Operators}
# Let's clear out our variables
rm(list=ls())

# Define the weights of some wolves (in pounds) around campus
wtFen <- 135 
wtArtemis <- 70 
wtAurora <- 70.000000001 
wtFang <- 45 

wtSnoopy <- 70 

# Let's compare Snoopy's weight to the other wolves

wtSnoopy < wtFen
wtSnoopy >= wtArtemis
wtSnoopy < wtAurora
wtSnoopy > wtFang

# Let's look at `==` more closely
wtSnoopy == wtArtemis # perfectly fine, two integers
wtSnoopy == wtAurora # Seems good because technically the weights are not equal
# BUT DON'T DO THIS

```

## Logical

```{r Logical-Operators}
# Use variables from last section

# Define some variables to compare
numWolves <- 50
numDogs <- 25
numCats <- 100

# Define some Boolean variables and set them
isParty <- FALSE
haveHotdogs <- TRUE
haveCheetos <- TRUE

# Check if numWolves is less than numDogs AND numCats is less than numDogs
(numWolves < numDogs) && (numCats < numDogs)

# Check if numDogs is less than or equal to 25 OR numCats is less than or equal to 25
(numDogs <= 25) || (numCats <= 25)

# Check if numDogs is less than 150 OR numCats is less than 150
(numDogs < 150) || (numCats < 150)

# See isParty is TRUE AND if numDogs is greater than 0 AND haveHotdogs is FALSE
# This could be a problem
isParty && (numDogs > 0) && !haveHotdogs

# See if (numDogs is greater than zero AND haveHotdogs is TRUE) OR (numCats is greater than 0 AND haveCheetos is TRUE)

(numDogs > 0 && haveHotdogs) || (numCats > 0 && haveCheetos)

# How could you fix the expression above to make sure that you have food available for both cats and dogs who come to your party?

(numDogs > 0 && haveHotdogs) && (numCats > 0 && haveCheetos)


```

